\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-17 rebuttal.tex

\documentclass{letter}
\usepackage[backend=biber]{biblatex}
\usepackage{todonotes}
\setuptodonotes{inline}

\addbibresource{references.bib}

\newcommand{\reply}[1]{%
	\textbf{Response to review comments made by prof. #1}

}

\begin{document}

\signature{
	Arne Van Den Kerchove \\
	\texttt{arne.vandenkerchove@kuleuven.be} \\
	+32 473 32 78 71 \\
}
\address{
	KU Leuven \\
	Laboratory for Neuro- and Psychophysiology \\
	Campus Gasthuisberg, O\&N 2 \\
	Herestraat 49 bus 1021 \\
	BE-3000 Leuven
}

\begin{letter}{
	prof. John Creemers \\
	KU Leuven \\
	Doctoral School of Biomedical Sciences \\
	Campus Gasthuisberg, O\&N 4 \\
	Herestraat 49 bus 700
	BE-3000 Leuven

}
\opening{Dear chair and members of the reading committee}

I would like to extend my sincere gratitude for time and effort you have taken
to review the manuscript of my PhD dissertation.
Additionally, I would like to thank you for the insightful, constructive and
detailed comments you all have provided.
The following reply contains a point-by-point response to your review comments.
Attached, you will also find an updated draft of the manuscript and a version
with indicated changes for your approval.

\reply{Andrea K\"ubler}
Major comments:
\begin{enumerate}
	\item
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item
\end{enumerate}

\reply{Fabien Lotte}
Major comments:
\begin{enumerate}
	\item Part of the confusion between active and passive BCI might be due
	to the poor choice of phrasing ``paradigms that require little active participation from the user, like redirect-
	ing attention or initiating imagined actions, are classified as passive BCI.''.
	This has been rewritten to ``in contrast to paradigms that require the user's active participation, for
	instance through redirecting attention or initiating imagined actions,
	paradigms free of this requirement are classified as \emph{passive \ac{bci}}''
	for clarity.
	Neurofeedback has been reclassified as active BCI in the text and in
	figure 1.3.
	\item \todo{Why XDAWN+RG no shrinkage (performed worse), fair
	comparison of sample rate}
	\item \todo{report actual accuracies performance stbf in table}
	\item In the results section of chapter 4, statistical comparisons were
	only performed between methods for which we evaluated the performance
	ourselves (HODA, PARAFACDA and BTTDA).
	Since the MOABB benchmarking framework~\cite{Aristimunha2023} allows for
	fair comparison of performance metrics reported across different works,
	the analyses for other comparison methods were not re-run
	by us as reported, but performances were
	taken from the MOABB benchmark database~\cite{Chevallier2024}.
	However, this source provided only the aggregated performance metrics
	and not the metrics per subject, session and cross-validation fold,
	hence we could not perform relevant statistical comparison with these
	comparison models.
	\item \todo{include table with absolute performances for transfer
	conditions}
\end{enumerate}
Minor comments:
\todo{minor comments in document}
\begin{enumerate}
	\item According to the annotations, typographical errors have been
	corrected and imprecise or inaccurate statements have been adjusted.
	The recommended references have been discussed and, where necessary,
	statements were adapted to accommodate this.
\end{enumerate}

\reply{Reinhold Scherer}
Major comments:
\begin{enumerate}
	\item \todo{quantify qualitative statements}
	\item  Mathematical equations and derivations have thoroughly
	been checked for correctness and notational consistency, and have been
	corrected throughout the manuscript where necessary.
	\item As indicated by the reviewer, true on-line experiments with live
	feedback have indeed not yet be conducted, but were foreseen to take
	place within the framework of this doctoral study.
	I agree that these experiments would be valuable to supplement the
	presented conclusions and are the logical next step.
	Indeed, the start of experimentation with healthy control participants
	as well as patient groups, was heavily delayed by the COVID-19 pandemic, as the
	preparatory research started in September 2019 and I officially
	enrolled in the the doctoral training programme in January 2020.
	This restricted not only our access to experiment subjects for a
	significant period at the start of the PhD, and even longer for
	vulnerable patient groups, but also shifted priorities for the medical
	staff at our partner institutions and patient centers.
	Hence, work early in the PhD mainly focused on the development of
	computational methods.
	We are currently starting up true on-line experimentation in cooperation
	with our partner patient centers, with an adapted interface and experimental protocol.
	\todo{indicate in manuscript covid}
	\item Recorded datasets could not be made publicly available since the
	ethical protocol (Ethics Commission of University Hospital Leuven --
	S62547 v1.4) for the conducted studies did not include provisions
	publishing the data.
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item
\end{enumerate}

\reply{Maarten De Vos}
Major comments:
\begin{enumerate}
	\item Mathematical equations and derivations have thoroughly
	been checked for correctness and notational consistency, and have been
	corrected throughout the manuscript where necessary.
	The paragraph on Block-Toeplitz linear discriminant analysis (tLDA) in
	section 1.6.4 has also been corrected, and it was expanded for clarity.
	Figures 3.1 and 3.2 have been regenerated and relabeled for consistent
	formatting.
	\item \todo{replace wilcoxon by delong}
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item Unsupported claims regarding generalisability have been omitted
	from the abstract.
	\item Throughout the manuscript, different EEG band-pass filtering
	settings have been used.
	Chapter 1 indicates that ERPs are usually band-pass filtered between
	0.5 and 16 Hz.
	These were also the preferred cut-off frequencies for other
	(unpublished) analyses used during the course of the thesis work, an
	those used for the analyses in chapter 3.
	However, ERP datasets evaluated in chapter 4 were band-pass filtered
	between 1 and 24 Hz, and in chapter 6 between 0.1 and 20 Hz.
	While there is no general consensus in the field about the optimal
	cut-off frequencies for P300 detection and ERP analysis in general,
	these values all fall well within the accepted
	bounds~\cite{Bougrain2012}.
	Decisions to deviate from this were made consciously.
	In chapter 3, filtering between 1 and 24 Hz was applied for consistency
	with the recommended ERP preprocessing pipeline of the MOABB
	benchmarking framework~\cite{Aristimunha2023}.
	This allows for a fair comparison under equal circumstances of the
	performance of the proposed algorithm with performances of other
	algorithms reported by the MOABB benchmarking
	database~\cite{Chevallier2024}.
	In chapter 6, the low-pass frequency was increased from 16 Hz to 20 Hz,
	since the analyses in this explicitly involve the short-time effects of
	latencies of ERP components, and the analysis of faster, early ERP
	components. A low low-pass frequency would harm the precision of
	latency effects in the filtered signal.
	A lower high-pass filter of 0.1 Hz was used here to reveal the
	influence of the proposed alignment approach on potential
	slower late components.
	\item Typographical errors have been corrected throughout the
	manuscript.
\end{enumerate}

\reply{Adalberto Simeone}
Major comments:
\begin{enumerate}
	\item \todo{discuss in limitations the impact of letters and future
	work experiments with an actual usable interface}
	\todo{mention here plans to go to patients with actual speller
	interface}
\end{enumerate}
Minor comments:
\begin{enumerate}
	\item \todo{correct si measurements Hz, mm, cm, V, µV, $\mu$ V}
	\item \todo{do proper check in textcite, especially in chapter 3}
	\item \todo{check for contractions and remove we've, isn't, wouldn't}
\end{enumerate}

\textbf{References}

\printbibliography[heading=none]

\closing{Kind regards and thank you in advance,}
\end{letter}
\end{document}

### 2024-11-17 chapters/bci.tex
Indicated comments Reinhold and Fabien

Starting in the 1970s from methods to establish basic communication from minimal brain
activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
into sophisticated interaction schemes that restore or enhance many capabilities by
bypassing defective parts of the nervous or muscle system.
By now, classic \acp{bci} methods are starting to mature outside the lab setting and can
be used as assistive communication technology by individuals with \acl{sspi}.
speech~\cite{Wolpaw2018,Milekovic2018}.
Impressive cutting-edge experimental examples include a brain-spine interface allowing a
paraplegic individual to walk again~\cite{Lorach2023}, a \ac{bci} translating internal
speech to a facial avatar speaking the imagined text~\cite{Metzger2023}, and fast
communication through decoding imagined handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
NeuraLink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

@@

Finally, the loop can be closed by coupling this to a device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact with
their environment.
For a communication system, this means coupling the decoded information to, e.g., a
virtual keyboard or speech synthesizer.
On the other hand, the actions performed by the \ac{bci} can themselves alter the user's
cognitive state or brain activity, creating an adaptive system.
This happens either directly through electrical neurostimulation, as is the
case for adaptive deep brain stimulation in Parkinson's disease or for the
cochlear implant as a hearing aide, or indirectly through sensory stimulation.
The latter involves specific, paradigm-related, sensory input, like tactile feedback for
movement \ac{bci}~\cite{Flesher2021}, or simply presenting selected actions back to the
user.
The user's brain will then adapt through reinforcement learning, causing changes in
behavior or strategy.
Neuroplasticity, the brain's ability to adapt and reinforce specific neural pathways,
can have a positive impact on \ac{bci} performance and gives rise to rehabilitation
applications.

@@

A better candidate is the neuronal electrical activity.
The brain contains 86 billion neurons, which are highly interconnected cells that are
the smallest units of information processing.
The \emph{action potential} and its related \emph{postsynaptic potential} are electrical
pulses occurring when a neuron receives input from another neuron and is activated.
The firing of a neuron, or the combined firing of groups of synchronized neurons, thus
generates an electrical field in and near the brain, which can be measured using
electrodes.
The related discipline is \emph{electrophysiology}.

@@

Specifically, invasive methods are less noisy and come with an increased \emph{spatial
resolution}, meaning they can extract a signal from a specific brain region or even a
set of neurons of interest.
This makes it easier to retrieve a signal with a known origin from the brain, which
might be leveraged by the paradigm to modulate information.
Generally, microelectrodes penetrating the cortex are considered to have the highest
spatial resolution.
These are needles with a diameter smaller than 10µm and one or multiple electrodes that
extend several millimeters into the cortex.
Microelectrodes can measure the Local Field Potential (LFP), the extracellular potential
of a small group of neurons, or even intracellular single neuron action potential
spikes.
They come in single-electrode form, or, more popularly, in microelectrode arrays.
Well-known examples are the Utah array~\cite{Maynard1997}, the more recent NeuraLink
implant~\cite{Musk2019}, and IMEC's Neuropixels 2.0~\cite{Steinmetz2021}.

Larger implanted electrodes are referred to as intracranial \ac{eeg} (i\ac{eeg}).
Depth electrodes, like those found in in stereo-\ac{eeg}, can be used for \ac{bci}~\cite{Wu2024}
 and other applications, like closed-loop adaptive control for Deep Brain Stimulation to mitigate Parkinson's disease
symptoms~\cite{Arlotti2018}.
Depth electrodes are longer rods with multiple electrodes that can extend deeper into
the brain.
Since these still penetrate the cortex, electrodes that sit on top of the cortex are
more popular, as they are considered less invasive.
\Ac{ecog} is a powerful method with many applications in \ac{bci}~\cite{Schalk2011} and
epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019}, with a higher
number of smaller electrodes per surface area, allow for more precise measurements.

@@

Paradigms that require little active participation from the user, like redirecting
attention or initiating imagined actions, are classified as \emph{passive \ac{bci}}.
From the user's point of view, this concept is probably the most attractive, since their
cognitive state would be inferred from just their brain activity.
However, without active user participation, it can be hard to establish reliable
communication or control.
Therefore, passive \ac{bci} are currently more suited to tasks like emotion or affect
monitoring~\cite{Torres2020,Libert2019, Muehl2014}, workload monitoring~\cite{Zanetti2021},
or, more dependent on stimulation, error detection~\cite{SiMohammed2020}.
It also has clinical applications in the form of neurofeedback~\cite{Hammond2011}.
Nevertheless, active participation facilitates training data collection for supervised
machine learning, since the conditions can be more easily controlled by the user or the
stimulation.

Paradigms requiring high active participation can be split up into \emph{active
\ac{bci}} and \emph{reactive \ac{bci}}.
Active \ac{bci} paradigms encode endogenous activity initiated by the user, such as
imagined movement or imagined speech.
These tasks can encode complex information, but current non-invasive communication
methods often limit the considered domain to a few movement directions to control a
cursor, or a few words~\cite{Panachakel2021}.
Invasive recordings are necessary for decoding more complex encoded activities like
natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
symbols~\cite{Willett2021} or sign language~\cite{Branco2017}.
Active paradigms are also subject to large inter-subject variability due to the
complexity of the performed tasks.
They might require extensive user training and are sensitive to correct task
instruction.
Imagined speech or movements can be performed in a multitude of ways, which can each
have different neural representations for separate individuals.
This can make it hard to adapt the \ac{bci} to specific individuals, causing poor
performance and giving rise to the concept of \ac{bci} illiteracy~\cite{Allison2010}\footnote{%
Recently, the concept of BCI illiteracy is increasingly being seen as outdated.
Instead of attributing poor performance to the user's inherent limitations, critiques
point out that the issue lies more with the design of the systems themselves.
BCI systems often fail to account for individual variability in brain anatomy and
cognitive strategies.
As a result, what has been labeled as `illiteracy' may, in fact, be a reflection of
systems that are not flexible enough to adapt to different users.
This perspective shifts the responsibility to the individual, rather than recognizing
that better, more adaptive systems could overcome many of these performance challenges,
thus standing in the way of progress.
Moreover, the term `illiteracy' itself carries negative connotations, implying a user
deficiency that reinforces normative assumptions, when in reality, a more inclusive
approach to design could enhance BCI accessibility for a wider range of
individuals.
This evolving view calls for a reframing of the problem, focusing on improving the
adaptability of BCIs rather than labeling users as illiterate~\cite{Becker2022,Thompson2019}.}.

@@

Together with \ac{ssvep}, oddball paradigms are most frequently used in visual \ac{bci}.
There are, however, indications that users prefer oddball stimulation over \ac{ssvep},
which can cause eye strain and fatigue due to the continuous oscillatory stimulation of
all targets at once~\cite{Xu2021}.
Furthermore, \ac{ssvep} relies more on directing the gaze at the intended target than
oddball, detracting from the concept of control independent from muscle movement.

@@

This leads to a problem that is not often addressed:
a visual \ac{bci} is dependent on the gaze of the user, and hence on muscle control.
This is exactly what \acp{bci} for augmented and assistive technology try to avoid.
The \ac{bci} target population consists of individuals with severe physical
impairment due to  with various degrees
of paralysis, or even with \ac{lis}, the complete loss of muscle control with preserved
consciousness.
While \ac{bci}s are most attractive as a solution for individuals with \ac{sspi},
studies paradoxically often report poor performance in exactly this group,
precisely due to their lack of adequate eye motor control.
Equally often, participants with gaze impairments were excluded from the study.

>>>>

Starting in the 1970s from methods to establish basic communication from minimal brain
activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
into sophisticated interaction schemes that restore or enhance many capabilities by
bypassing defective parts of the nervous or muscle system.\todo{the BCI concetp
was proposed at that time, but maybe not yet this form of communication. Also,
this reference is not from the 1970s, you may want to cite the original paper}
By now, classic \acp{bci} methods are starting to mature outside the lab setting and can
be used as assistive communication technology by individuals with \acl{sspi}
~\cite{Wolpaw2018,Milekovic2018}.
Impressive cutting-edge experimental examples include a brain-spine interface allowing a
paraplegic individual to walk again~\cite{Lorach2023}, a \ac{bci} translating internal
speech to a facial avatar speaking the imagined text~\cite{Metzger2023}, and fast
communication through decoding imagined handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
NeuraLink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

@@

Finally, the loop can be closed by coupling this to a device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact with
their environment.
For a communication system, this means coupling the decoded information to, e.g., a
virtual keyboard or speech synthesizer.
On the other hand, the actions performed by the \ac{bci} can themselves alter the user's
cognitive state or brain activity, creating an adaptive system.
This happens either directly through electrical neurostimulation, as is the
case for adaptive deep brain stimulation in Parkinson's disease or for the
cochlear implant as a hearing aide, or indirectly through sensory stimulation.
The latter involves specific, paradigm-related, sensory input, like tactile feedback for
movement \ac{bci}~\cite{Flesher2021}, or simply presenting selected actions back to the
user.
The user's brain will then adapt through learning, causing changes in
behavior or strategy.
Neuroplasticity, the brain's ability to adapt and reinforce specific neural pathways,
can have a positive impact on \ac{bci} performance and gives rise to rehabilitation
applications.

@@

A better candidate is the neuronal electrical activity.
The brain contains 86 billion neurons, which are highly interconnected cells that are
the smallest units of information processing.
The \emph{action potential} and its related \emph{postsynaptic potential} are electrical
pulses occurring when a neuron receives input from other neurons and is activated.
The firing of a neuron, or the combined firing of groups of synchronized neurons, thus
generates an electrical field in and near the brain, which can be measured using
electrodes.
The related discipline is \emph{electrophysiology}.

@@

Specifically, invasive methods are less noisy and come with an increased \emph{spatial
resolution}, meaning they can extract a signal from a specific brain region or even a
set of neurons of interest.
This makes it easier to retrieve a signal with a known origin from the brain, which
might be leveraged by the paradigm to modulate information.
Generally, microelectrodes penetrating the cortex are considered to have the highest
spatial resolution.
These are needles with a diameter smaller than 10 µm and one or multiple electrodes that
extend several millimeters into the cortex.
Microelectrodes can measure the Local Field Potential (LFP), the extracellular potential
of a small group of neurons,\todo{qualitative}
or even intracellular single neuron action potential
spikes.
They come in single-electrode form, or, more popularly, in microelectrode arrays.
Well-known examples are the Utah array~\cite{Maynard1997}, the more recent NeuraLink
implant~\cite{Musk2019}, and IMEC's Neuropixels 2.0~\cite{Steinmetz2021}.

Larger implanted electrodes are referred to as intracranial \ac{eeg} (i\ac{eeg}).
Depth electrodes, like those found in in stereo-\ac{eeg}, can be used for \ac{bci}~\cite{Wu2024}
 and other applications, like closed-loop adaptive control for Deep Brain Stimulation to mitigate Parkinson's disease
symptoms~\cite{Arlotti2018}.\todo{qualify size}
Depth electrodes are longer rods with multiple electrodes that can extend deeper into
the brain.
Since these still penetrate the cortex, electrodes that sit on top of the cortex are
more popular, as they are considered less invasive.
\Ac{ecog} is a powerful method with many applications in \ac{bci}~\cite{Schalk2011} and
epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG), with a higher
number of smaller electrodes per surface area (recording from sites with
diameter of 10s of micrometers instead of $\pm$ 1 cm)~\cite{Shokoueinejad2019},
allow for more precise measurements.

@@

In contrast to paradigms that require the user's active participation, for
instance through redirecting attention or initiating imagined actions,
paradigms free of this requirement are classified as \emph{passive \ac{bci}}.
From the user's point of view, this concept is probably the most attractive, since their
cognitive state would be inferred from just their brain activity.
However, without active user participation, it can be hard to establish reliable
communication or control.
Therefore, passive \ac{bci} are currently more suited to tasks like emotion or affect
monitoring~\cite{Torres2020,Libert2019, Muehl2014}, workload monitoring~\cite{Zanetti2021},
or, more dependent on stimulation, error detection~\cite{SiMohammed2020}.
Clinical applications are found in the diagnosis of a multitude of neurological
and psychological disorders through various forms of neuroimaging, in resting
state or while performing a specific task.
Nevertheless, active participation facilitates training data collection for supervised
machine learning, since the conditions can be more easily controlled by the user or the
stimulation.

Paradigms that do require high active participation can be split up into \emph{active
\ac{bci}} and \emph{reactive \ac{bci}}.
Active \ac{bci} paradigms encode endogenous activity initiated by the user, such as
imagined movement or imagined speech.
These tasks can encode complex information, but current non-invasive communication
methods often limit the considered domain to a few movement directions to control a
cursor, or a few words~\cite{Panachakel2021}.
Invasive recordings are necessary for decoding more complex encoded activities like
natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
symbols~\cite{Willett2021} or sign language~\cite{Branco2017}.
Active BCI also has clinical applications in the form of
neurofeedback~\cite{Hammond2011}, where the user adapts mental strategies to
perform active self-regulation of their brain activity which is presented to
them, causing a feedback loop.

Active paradigms are subject to large inter-subject variability due to the
complexity of the performed tasks.
They might require extensive user training and are sensitive to correct task
instruction.
Imagined speech or movements, or mental strategies in general, can be
performed in a multitude of ways, which can themselves have different neural
representations for separate individuals.
This can make it hard to adapt the \ac{bci} to specific individuals, causing poor
performance and giving rise to the concept of \ac{bci} illiteracy~\cite{Allison2010}\footnote{%
Recently, the concept of BCI illiteracy is increasingly being seen as outdated.
Instead of attributing poor performance to the user's inherent limitations, critiques
point out that the issue lies more with the design of the systems themselves.
BCI systems often fail to account for individual variability in brain anatomy and
cognitive strategies.
As a result, what has been labeled as `illiteracy' may, in fact, be a reflection of
systems that are not flexible enough to adapt to different users.
This perspective shifts the responsibility to the individual, rather than recognizing
that better, more adaptive systems could overcome many of these performance challenges,
thus standing in the way of progress.
Moreover, the term `illiteracy' itself carries negative connotations, implying a user
deficiency that reinforces normative assumptions, when in reality, a more inclusive
approach to design could enhance BCI accessibility for a wider range of
individuals.
This evolving view calls for a reframing of the problem, focusing on improving the
adaptability of BCIs rather than labeling users as illiterate~\cite{Becker2022,Thompson2019}.}.

@@

Together with \ac{ssvep}, oddball paradigms are most frequently used in visual \ac{bci}.
There are, however, indications that users prefer oddball stimulation over \ac{ssvep},
which can cause eye strain and fatigue due to the continuous oscillatory stimulation of
all targets at once~\cite{Xu2021}\footnote{
  Recent work mitigates this by making SSVEP stimulation
  more comfortable without compormising selection accuracy through
  high-frequency low-contrast stimulation~\cite{Ladouce2022}.
}.
Furthermore, \ac{ssvep} relies more on directing the gaze at the intended target than
oddball, detracting from the concept of control independent from muscle movement.

@@

This leads to a problem that is not often addressed:
a visual \ac{bci} is dependent on the gaze of the user, and hence on muscle control.
This is exactly what \acp{bci} for augmented and assistive technology try to avoid.
The \ac{bci} target population consists of individuals with severe physical
impairment due to  with various degrees
of paralysis, or even with \ac{lis}, the complete loss of muscle control with preserved
consciousness.
While \ac{bci}s are most attractive as a solution for individuals with \ac{sspi},
studies paradoxically often report poor performance in exactly this group.
This is caused by a multitude of factors, including diminished
electrophysiological responses and cognitive defects linked to the
pathophysiology, psychological factors such as the ``extinction of goal directed thinking''
phenomenon in complete impairment, and, importantly, sensory impairments such as vision and
oculomotor defects~\cite{Seguin2023}.
While visual \acp{bci} often rely on visual sensory functions, it is precisely this
group of patients that could benefit from them, that lacks adequate eye motor control.
Equally often, participants with gaze impairments were excluded from the study.
